% !TEX root = ../Thesis.tex
\chapter{Usage of AI}

% TODO: Used AI tools, like ChatGPT, to help with the writing of this thesis. I used it to generate ideas, to write parts of the text, to improve the text, and to check the grammar and spelling. I also used it to generate code snippets and to explain code snippets.

For the development of this thesis, AI-assisted technologies, specifically large language models, were utilized to enhance various aspects of the writing and research process.

\begin{itemize}
    \item \textbf{Text Transformation and Fluency:} AI tools were primarily used to refine and transform sections of the text to improve fluency, clarity, and highlight important aspects without altering the original content or technical accuracy. This included rephrasing sentences, improving sentence structure, and ensuring a consistent academic tone.
    \item \textbf{Idea Generation and Structuring:} In the initial phases, AI was employed to brainstorm ideas for different chapters, structure the thesis content logically, and expand on key concepts.
    \item \textbf{Grammar and Spelling Checks:} AI tools assisted in reviewing the thesis for grammatical errors, spelling mistakes, and punctuation issues, contributing to the overall linguistic quality of the document.
    \item \textbf{Code Snippet Assistance:} AI was also used to generate and explain small code snippets, which aided in understanding certain programming constructs or illustrating concepts within the practical implementation sections.
\end{itemize}

It is important to note that while AI provided significant assistance, the core research, conceptual design, implementation, and analytical interpretation remained the sole responsibility of the author. All information presented in this thesis, including any text passages or code generated with AI assistance, has been thoroughly reviewed, verified, and integrated by the author to ensure accuracy, originality, and adherence to academic standards.
